---
title: "Moose Landcape Change Hypothesis Analysis"
author: "Tyler Muhly"
date: "12/12/2019"
output: 
  html_document:
  keep_md: yes
---
<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (data.table)
library (downloader)
library (dplyr)
library (ggplot2)
library (here)
library (lwgeom)
library (sf)
library (kableExtra)
```

## Introduction
Here I summarize data to explore the moose 'landscape change hypothesis' (LCH). The moose LCH posits that recent moose population declines in parts of central British Columbia (BC) were related to large-scale changes to forest cover (i.e., landscape disturbance), caused by an extensive mountain pine beetle (MBP) infestation and subsequent forest harvest to salvage MPB killed forest stands. As a preliminary test of the LCH, I relate estimates of forest disturbance from MPB and forestry (cutblocks) to estimates of adult female moose survival and cow/calf moose ratios measured in five study areas. 

The LCH predicts a negative correlative relationship between the amount of landscape disturbance and adult female survival or the ratio of calves to cow moose. This relationship is indicative that processes related to landscape disturbance, i.e., reduced amount of security and thermal cover and increased amount of negative interactions with humans (e.g., hunting mortality) due to higher road densities are causing moose population declines. 

## Methods
### Study Areas
The LCH was tested by sampling moose survival rates in the Big Creek, Bonaparte, Entiako, John Prince Research Forest (JPRF) and Prince George South study areas. These five study areas were selected because they had varied levels of MPB and forestry disturbance. A sample of moose in each study area were captured annually and collared with global positioning system (GPS) telemetry collars. Annual moose study area boundaries were determined by calculating the 100% minimum convex polygon (MCP) of annual sampled adult female moose GPS locations in each area. Moose locations were collected for four years from 2015 to 2018 (except Entiako, where data was also collected in 2014), and thus study area boundaries were re-calculated for each of these years.  

### Population Data
Adult female moose survival rates were estimated annually from mortality data collected from the sampled animals. Survival was estimated for a biological year, i.e., April to April(?), not a calendar year. Thus, for example, a survival estimate for 2015 was caculated using mortality data collected from April 2014 to April 2015. In addition, aerial cow/calf composition surveys were completed annually in each study area in ????.

```{r, management unit spatial data, eval = F, echo = F, message = F, warning = F, include = F}
# load the mcp's
poly.sa.bc.14.15 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "BigCreek_2014_15_MCP"), 
                         crs = 3005) # setting these all to the same coordinate system
poly.sa.bc.14.15$year <- 2015 # add year
poly.sa.bc.15.16 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "BigCreek_2015_16_MCP"), 
                         crs = 3005)
poly.sa.bc.15.16$year <- 2016 # add year
poly.sa.bc.16.17 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "BigCreek_2016_17_MCP"), 
                         crs = 3005)
poly.sa.bc.16.17$year <- 2017 # add year
poly.sa.bc.17.18 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "BigCreek_2017_18_MCP"), 
                         crs = 3005)
poly.sa.bc.17.18$year <- 2018 # add year
poly.sa.jprf.14.15 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "JPRF_2014_15_MCP"), 
                         crs = 3005)
poly.sa.jprf.14.15$year <- 2015 
poly.sa.jprf.15.16 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "JPRF_2015_16_MCP"), 
                         crs = 3005)
poly.sa.jprf.15.16$year <- 2016 
poly.sa.jprf.16.17 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "JPRF_2016_17_MCP"), 
                         crs = 3005)
poly.sa.jprf.16.17$year <- 2017
poly.sa.jprf.17.18 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "JPRF_2017_18_MCP"), 
                         crs = 3005)
poly.sa.jprf.17.18$year <- 2018
poly.sa.pgs.14.15 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "PGS_2014_15_MCP"), 
                         crs = 3005)
poly.sa.pgs.14.15$year <- 2015
poly.sa.pgs.15.16 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "PGS_2015_16_MCP"), 
                         crs = 3005)
poly.sa.pgs.15.16$year <- 2016
poly.sa.pgs.16.17 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "PGS_2016_17_MCP"), 
                         crs = 3005)
poly.sa.pgs.16.17$year <- 2017
poly.sa.pgs.17.18 <- st_transform (st_read (dsn = here ( "./data/mcp/PGS_JPRF_BigCreek_MCPs.gdb/PGS_JPRF_BigCreek_MCPs.gdb"),
                         layer = "PGS_2017_18_MCP"), 
                         crs = 3005)  
poly.sa.pgs.17.18$year <- 2018
poly.sa.bon.14.15 <- st_transform (st_read (dsn = here ( "./data/mcp/bonaparte/Bon14_15MCP.shp")), 
                         crs = 3005)
poly.sa.bon.14.15$year <- 2015
poly.sa.bon.15.16 <- st_transform (st_read (dsn = here ( "./data/mcp/bonaparte/Bon15_16MCP.shp")), 
                         crs = 3005)
poly.sa.bon.15.16$year <- 2016
poly.sa.bon.16.17 <- st_transform (st_read (dsn = here ( "./data/mcp/bonaparte/Bon16_17MCP.shp")), 
                         crs = 3005)
poly.sa.bon.16.17$year <- 2017
poly.sa.bon.17.18 <- st_transform (st_read (dsn = here ( "./data/mcp/bonaparte/Bon17_18MCP.shp")), 
                         crs = 3005)
poly.sa.bon.17.18$year <- 2018
poly.sa.entiako <- st_transform (st_read (dsn = here ( "./data/mcp/entiako/mmcp100.shp")), 
                         crs = 3005)

# bind them up by area, add area names adn make fields consistent
# Big Creek
poly.sa.bc <- rbind (poly.sa.bc.14.15, poly.sa.bc.15.16, poly.sa.bc.16.17, poly.sa.bc.17.18)
poly.sa.bc$name_sa <- "Big Creek"
poly.sa.bc$area_ha <- poly.sa.bc$area * 0.0001
poly.sa.bc$Shape_Length <- NULL
poly.sa.bc$Shape_Area <- NULL
poly.sa.bc$area <- NULL
poly.sa.bc$id <- 1
names(poly.sa.bc)[names(poly.sa.bc) == "Shape"] = "geometry" # rename the geometry column
st_geometry (poly.sa.bc) <- "geometry"
rm (poly.sa.bc.14.15, poly.sa.bc.15.16, poly.sa.bc.16.17, poly.sa.bc.17.18)
gc ()

# John Prince Research Forest
poly.sa.jprf <- rbind (poly.sa.jprf.14.15, poly.sa.jprf.15.16, poly.sa.jprf.16.17, poly.sa.jprf.17.18)
names(poly.sa.jprf)[names(poly.sa.jprf) == "Id"] = "id" # rename the Id column
poly.sa.jprf$id <- 2
poly.sa.jprf$area_ha <- (st_area (poly.sa.jprf$Shape) * 0.0001) # area needs to be calc'd
poly.sa.jprf$name_sa <- "John Prince Research Forest"
poly.sa.jprf$Shape_Length <- NULL
poly.sa.jprf$Shape_Area <- NULL
names(poly.sa.jprf)[names(poly.sa.jprf) == "Shape"] = "geometry" # rename the geometry column
st_geometry (poly.sa.jprf) <- "geometry"
rm (poly.sa.jprf.14.15, poly.sa.jprf.15.16, poly.sa.jprf.16.17, poly.sa.jprf.17.18)
gc ()

# Prince George South
poly.sa.pgs <- rbind (poly.sa.pgs.14.15, poly.sa.pgs.15.16, poly.sa.pgs.16.17, poly.sa.pgs.17.18)
names(poly.sa.pgs)[names(poly.sa.pgs) == "Id"] = "id" 
poly.sa.pgs$id <- 3
poly.sa.pgs$area_ha <- (st_area (poly.sa.pgs$Shape) * 0.0001)
poly.sa.pgs$name_sa <- "Prince George South"
poly.sa.pgs$Shape_Length <- NULL
poly.sa.pgs$Shape_Area <- NULL
names(poly.sa.pgs)[names(poly.sa.pgs) == "Shape"] = "geometry" # rename the geometry column
st_geometry (poly.sa.pgs) <- "geometry"
rm (poly.sa.pgs.14.15, poly.sa.pgs.15.16, poly.sa.pgs.16.17, poly.sa.pgs.17.18)
gc ()

# Bonaparte
poly.sa.bon <- rbind (poly.sa.bon.14.15, poly.sa.bon.15.16, poly.sa.bon.16.17, poly.sa.bon.17.18)
names(poly.sa.bon)[names(poly.sa.bon) == "Id"] = "id" 
poly.sa.bon$id <- 4
poly.sa.bon$area_ha <- (st_area (poly.sa.bon$geometry) * 0.0001)
poly.sa.bon$name_sa <- "Bonaparte"
rm (poly.sa.bon.14.15, poly.sa.bon.15.16, poly.sa.bon.16.17, poly.sa.bon.17.18)
gc ()

# Entiako
poly.sa.entiako$year <- 2014
poly.sa.entiako [2, 4] <- 2015 # making year consitent with other data
poly.sa.entiako [3, 4] <- 2016
poly.sa.entiako [4, 4] <- 2017
poly.sa.entiako [5, 4] <- 2018
poly.sa.entiako$name_sa <- "Entiako"
poly.sa.entiako$id <- 5
poly.sa.entiako$area_ha <- poly.sa.entiako$area
poly.sa.entiako$area <- NULL

# all together now...
poly.sa <- rbind (poly.sa.bc, poly.sa.jprf, poly.sa.pgs, poly.sa.bon, poly.sa.entiako)
rm (poly.sa.bc, poly.sa.jprf, poly.sa.pgs, poly.sa.bon, poly.sa.entiako)
gc ()

# unique identifier for each
poly.sa$unique_id <- with (poly.sa, paste0 (as.character (name_sa), as.character (year)))

# save it 
st_write (poly.sa, 
          dsn = here ("./data/mcp/"), 
          layer = "study_area_mcp_moose", 
          driver = "ESRI Shapefile")
```

### Estimating Area of Cutblocks in Study Areas
I estimated the area of less than 8 and less than 15 year old forestry cutblocks and MPB disturbed stands in each study area. I estimated the area of cutblocks using the [consolidated cutblocks](https://catalogue.data.gov.bc.ca/dataset/harvested-areas-of-bc-consolidated-cutblocks-) dataset for BC. I intersected this data with the study area boundaries and then summarized the area cut by year from 1970 to 2018. I then calculated the area of cutblocks less than 8 and 15 years old for each study area year to align with adult female moose survival estimates for that year. For example, for survival estimates calculated for year 2015, I calculated the area cut from 2008 to 2015 as an estimate of area cut less than 8 years old. Thus, the assumption was the majority of area was cut early in a calendar year, during the period which survival estimates were calculated. 

```{r, cutblock data download and attribute to study areas, eval = F, echo = T, message = F, warning = F, include = F}

# library (bcdata) # use this to conenct to bcgw via R
# cutblocks <- bcdc_get_data ('b1b647a6-f271-42e0-9cd0-89ec24bce9f7', resource = '66656966-8af6-4e32-8de1-b79c63cae40f') # name of the data in BCGW; I noticed there are 413451 rows form here, whereas 446,987 in the BCGW; not sure why they are different, but this version likely underestimates the area of cutblocks; so I used the BCGW data and imported it here...
cutblocks <- sf::st_read (dsn = "C:\\Work\\caribou\\clus_data\\moose\\moose_lch.gdb",
                          layer = "cutblocks_20191217")
cutblocks <- lwgeom::st_make_valid (cutblocks) # fix any geometry errors

# intersect cutblocks with study area values
sa.cutblocks <- sf::st_intersection (cutblocks, poly.sa ["unique_id"]) # intersect the study area ID
table.sa.cutblocks <- as.data.table (sa.cutblocks)

# summarize cut by study area
table.sa.cutblocks.mum <- table.sa.cutblocks %>% 
                             group_by (unique_id, HARVEST_YEAR) %>%
                             summarise (area_ha = sum (AREA_HA)) # calculate area of cut by year

# get study area id
poly.sa.habitat <- as.data.table (poly.sa)
poly.sa.habitat$geometry <- NULL

## these next steps left-join the cutblock data to wmu data ##
### Subset the harvest years ###
#### 1980 to 2018 ####
fxn.cut.year <- function (year, table.sa.cutblocks.mum, poly.sa.habitat) {
  table <- as.data.table (table.sa.cutblocks.mum [table.sa.cutblocks.mum$HARVEST_YEAR == year, ])
  table [, paste0 ("cut_", year, "_area_ha") := area_ha] 
  table$area_ha <- NULL
  table$HARVEST_YEAR <- NULL
  setkey (table, unique_id)
  setkey (poly.sa.habitat, unique_id)
  poly.sa.habitat <- merge (poly.sa.habitat, table, all.x = TRUE)
  poly.sa.habitat [is.na (poly.sa.habitat)] <- 0 # if NA values
  poly.sa.habitat <<- poly.sa.habitat
}

year <- 2018
fxn.cut.year (year, table.sa.cutblocks.mum, poly.sa.habitat)
  
#### 1970 to 1979 ####
table.sa.cutblocks.1970to1979.mum <- table.sa.cutblocks.mum [table.sa.cutblocks.mum$HARVEST_YEAR > 1969 & table.sa.cutblocks.mum$HARVEST_YEAR < 1980, ]
table.sa.cutblocks.1970to1979.mum <- table.sa.cutblocks.1970to1979.mum %>% 
                                          group_by (unique_id) %>%
                                          summarise (area_ha = sum (area_ha))
table.sa.cutblocks.1970to1979.mum$cut_1970to1979_area_ha <- table.sa.cutblocks.1970to1979.mum$area_ha
table.sa.cutblocks.1970to1979.mum$area_ha <- NULL
table.sa.cutblocks.1970to1979.mum <- as.data.table(table.sa.cutblocks.1970to1979.mum)
setkey (table.sa.cutblocks.1970to1979.mum, unique_id)
setkey (poly.sa.habitat, unique_id)
poly.sa.habitat <- merge (poly.sa.habitat, table.sa.cutblocks.1970to1979.mum, all.x = TRUE)
poly.sa.habitat [is.na (poly.sa.habitat)] <- 0 # if NA values
rm (table.sa.cutblocks.1970to1979.mum)
gc ()

# 8 year cumulative cut disturbance 
poly.sa.habitat [year == 2014, cut_8_year_area_ha := cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha]
poly.sa.habitat [year == 2015, cut_8_year_area_ha := cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha] #where year = 2015
poly.sa.habitat [year == 2016, cut_8_year_area_ha := cut_2016_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha] #where year = 2016
poly.sa.habitat [year == 2017, cut_8_year_area_ha := cut_2017_area_ha + cut_2017_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha] 
poly.sa.habitat [year == 2018, cut_8_year_area_ha := cut_2018_area_ha + cut_2017_area_ha + cut_2017_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha] 

# 15 year cumulative cut disturbance 
poly.sa.habitat [year == 2014, cut_25_year_area_ha := cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha + cut_2006_area_ha + cut_2005_area_ha + cut_2004_area_ha + cut_2003_area_ha + cut_2002_area_ha + cut_2001_area_ha + cut_2000_area_ha + cut_1999_area_ha + cut_1998_area_ha + cut_1997_area_ha + cut_1996_area_ha + cut_1995_area_ha + cut_1994_area_ha + cut_1993_area_ha + cut_1992_area_ha + cut_1991_area_ha + cut_1990_area_ha]
poly.sa.habitat [year == 2015, cut_25_year_area_ha := cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha + cut_2006_area_ha + cut_2005_area_ha + cut_2004_area_ha + cut_2003_area_ha + cut_2002_area_ha + cut_2001_area_ha + cut_2000_area_ha + cut_1999_area_ha + cut_1998_area_ha + cut_1997_area_ha + cut_1996_area_ha + cut_1995_area_ha + cut_1994_area_ha + cut_1993_area_ha + cut_1992_area_ha + cut_1991_area_ha] 
poly.sa.habitat [year == 2017, cut_25_year_area_ha := cut_2017_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha + cut_2006_area_ha + cut_2005_area_ha + cut_2004_area_ha + cut_2003_area_ha + cut_2002_area_ha + cut_2001_area_ha + cut_2000_area_ha + cut_1999_area_ha + cut_1998_area_ha + cut_1997_area_ha + cut_1996_area_ha + cut_1995_area_ha + cut_1994_area_ha + cut_1993_area_ha + cut_1992_area_ha] 
poly.sa.habitat [year == 2017, cut_25_year_area_ha := cut_2017_area_ha + cut_2017_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha + cut_2006_area_ha + cut_2005_area_ha + cut_2004_area_ha + cut_2003_area_ha + cut_2002_area_ha + cut_2001_area_ha + cut_2000_area_ha + cut_1999_area_ha + cut_1998_area_ha + cut_1997_area_ha + cut_1996_area_ha + cut_1995_area_ha + cut_1994_area_ha + cut_1993_area_ha] 
poly.sa.habitat [year == 2018, cut_25_year_area_ha := cut_2018_area_ha + cut_2017_area_ha + cut_2017_area_ha + cut_2015_area_ha + cut_2014_area_ha + cut_2013_area_ha + cut_2012_area_ha + cut_2011_area_ha + cut_2010_area_ha + cut_2009_area_ha + cut_2008_area_ha + cut_2007_area_ha + cut_2006_area_ha + cut_2005_area_ha + cut_2004_area_ha + cut_2003_area_ha + cut_2002_area_ha + cut_2001_area_ha + cut_2000_area_ha + cut_1999_area_ha + cut_1998_area_ha + cut_1997_area_ha + cut_1996_area_ha + cut_1995_area_ha + cut_1994_area_ha] 

# save the data
write.csv (poly.sa.habitat, here ("./data/study_area_habitat.csv"))

rm (cutblocks, table.sa.cutblocks, table.sa.cutblocks.mum, sa.cutblocks)
gc ()
```

### Estimating Area of Mountain Pine Beetle Disturbed Stands in Study Areas
I estimated the area of forest stands disturbed by mountain pine beetle in each study area using data collected from [annual aerial surveys](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/forest-health/aerial-overview-surveys/methods) by the government of BC. These data were accessible through an [FTP site](https://www.for.gov.bc.ca/ftp/HFP/external/!publish/Aerial_Overview/), and I downloaded annual datasets from 1999 to 2018.  I summarized the area of moderately (11 to 29%), severely (30 to 49%) and very severely (>50%) disturbed forest stands for each year. Estimates cannot be added cumulatively each year, as damage to stands measured in previous years may be updated with new data from subsequent years. Thus I ensured that spatially overlapping stands were updated using ArcGIS. I then calculated the area of MPB disturbed stands less than 8 and less than 15 years old for each study area year to align with adult female moose survival estimates for that year. MPB surveys were typically completed in July-August, thus, for example, for survival estimates calculated for year 2015, I calculated the area of MPB disturbed stands from 2008 to 2015 as an estimate of area of MPB stands less than 8 years old. 

```{r, forest health data download and attribute beetle disturbance to study areas, echo = T, message = F, warning = F, eval = F, include = F}

# Download data; this is an example; do these for each year and save it locally
download ("https://www.for.gov.bc.ca/ftp/HFP/external/!publish/Aerial_Overview/1999/shape/Prov_Shape.zip", 
            dest = "C:\\Work\\caribou\\clus_data\\forest_health\\1999\\shape.zip", 
            mode = "wb")
unzip ("C:\\Work\\caribou\\clus_data\\forest_health\\1999\\shape.zip", 
       exdir = "C:\\Work\\caribou\\clus_data\\forest_health\\1999")
file.remove ("C:\\Work\\caribou\\clus_data\\forest_health\\1999\\shape.zip")

#########################################################
# Determine MPB affected stands by year - 1998 to 2008 #
#######################################################
#------------------------------------------------------
# I found that MPB data wasn't updated annualy, e.g., a surveyed stand in 1998 
# may not necessarily be re-surveyd in 1999. So, I had to 'erase' current years data from previous years data (cumulative), and then merge that with the cumulative mpb disturbance to date
### in R, but it TOOK AWHILE, SO I DID THE GIS PROCESSING IN ARCGIS ###

# do the analysis using the data processed in ArcGIS
  # clipped to study area
  # unioned with study area
  # area in ha calc (POLY_AREA)
  # erase new data from previous data
  # merge new data to cumulative data

# function to load and overlay mpb data by year and study area 

fxn.mpb.summary <- function (dsn, survey.year, poly.sa.habitat) {
fh <- sf::st_read (dsn = paste0 (dsn))
mpb <- fh %>%
        filter(FHF == "IBM") %>%
        filter (SEVERITY == "M" | SEVERITY == "S" | SEVERITY == "V")
table <- as.data.table (mpb)  
table$survey.year <- paste0 (survey.year)
table.cast <- as.data.table (dcast (table, 
                                    unique_id ~ YEAR_1 + SEVERITY + survey.year, 
                                    value.var = "POLY_AREA",  
                                    fun = sum))
setkey (table.cast, unique_id)
setkey (poly.sa.habitat, unique_id)
poly.sa.habitat <- merge (poly.sa.habitat, table.cast, all.x = TRUE)
poly.sa.habitat [is.na (poly.sa.habitat)] <- 0
poly.sa.habitat <<- poly.sa.habitat
}

fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_1999_sa_union.shp", 
 "survey1999", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2000_cumul.shp", 
  "survey2000", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2001_cumul.shp", 
  "survey2001", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2002_cumul.shp", 
 "survey2002", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2003_cumul.shp", 
 "survey2003", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2004_cumul.shp", 
  "survey2004", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2005_cumul.shp", 
  "survey2005", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2006_cumul.shp", 
  "survey2006", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2007_cumul.shp", 
  "survey2007", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2008_cumul.shp", 
  "survey2008", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2009_cumul.shp", 
  "survey2009", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2010_cumul.shp", 
  "survey2010", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2011_cumul.shp", 
  "survey2011", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2012_cumul.shp", 
  "survey2012", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2013_cumul.shp", 
  "survey2013", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2014_cumul.shp", 
 "survey2014", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2015_cumul.shp", 
  "survey2015", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2016_cumul.shp", 
  "survey2016", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2017_cumul.shp", 
  "survey2017", 
   poly.sa.habitat)
fxn.mpb.summary (
  "C:\\Work\\caribou\\clus_data\\forest_health\\mpb_sa\\fh_2018_cumul.shp", 
  "survey2018", 
   poly.sa.habitat)

# severity types: very severe (>50% of Trees in Polygon Recently Killed), severe (30% to 49%), moderate (11% to 29%)

# Cumulative mpb  
### NOTE: for MPB, each year survey is cumulative, impact of previous years are included in last year, so use last survey year 
poly.sa.habitat [year == 2014, mpb_m_8_year_area_ha_2014 := `X2014_M_survey2014` + `X2013_M_survey2014` + `X2012_M_survey2014` + `X2011_M_survey2014` + `X2010_M_survey2014` + `X2009_M_survey2014` + `X2008_M_survey2014` + `X2007_M_survey2014`]
poly.sa.habitat [year == 2014, mpb_s_8_year_area_ha_2014 :=    `X2012_S_survey2014` + `X2010_S_survey2014` + `X2009_S_survey2014` + `X2008_S_survey2014` + `X2007_S_survey2014` ] # no data   `X2014_S_survey2014` +  `X2013_S_survey2014` + `X2011_S_survey2014`  + 
poly.sa.habitat [year == 2014, mpb_vs_8_year_area_ha_2014 :=    `X2011_V_survey2014` +   `X2009_V_survey2014` + `X2008_V_survey2014` + `X2007_V_survey2014`] # no data   `X2014_V_survey2014` + `X2013_V_survey2014` + `X2012_V_survey2014` + `X2010_V_survey2014` +

poly.sa.habitat [year == 2015, mpb_m_8_year_area_ha_2015 := `X2015_M_survey2015` + `X2014_M_survey2015` + `X2013_M_survey2015` + `X2012_M_survey2015` + `X2011_M_survey2015` + `X2010_M_survey2015` + `X2009_M_survey2015` + `X2008_M_survey2015`]
poly.sa.habitat [year == 2015, mpb_s_8_year_area_ha_2015 :=  `X2012_S_survey2015` + `X2010_S_survey2015` + `X2009_S_survey2015` + `X2008_S_survey2015`] # no data `X2015_S_survey2015` +  `X2014_S_survey2015` + `X2013_S_survey2015` + `X2011_S_survey2015` + 
poly.sa.habitat [year == 2015, mpb_vs_8_year_area_ha_2015 :=    `X2011_V_survey2015` +  `X2009_V_survey2015` + `X2008_V_survey2015`] # no data `X2015_V_survey2015` +  `X2014_V_survey2015` +  `X2013_V_survey2015` + `X2012_V_survey2015` + `X2010_V_survey2015` + 

poly.sa.habitat [year == 2016, mpb_m_8_year_area_ha_2016 := `X2016_M_survey2016` + `X2015_M_survey2016` + `X2014_M_survey2016` + `X2013_M_survey2016` + `X2012_M_survey2016` + `X2011_M_survey2016` + `X2010_M_survey2016` + `X2009_M_survey2016`]
poly.sa.habitat [year == 2016, mpb_s_8_year_area_ha_2016 :=  `X2016_S_survey2016` +   `X2012_S_survey2016` +  `X2010_S_survey2016` + `X2009_S_survey2016`] # no data  `X2015_S_survey2016` + `X2014_S_survey2016` + `X2013_S_survey2016` +  `X2011_S_survey2016` +
poly.sa.habitat [year == 2016, mpb_vs_8_year_area_ha_2016 := `X2011_V_survey2016` +  `X2009_V_survey2016`] # no data  `X2016_V_survey2016` + `X2015_V_survey2016` + `X2014_V_survey2016` + `X2013_V_survey2016` +  `X2012_V_survey2016` + `X2010_V_survey2016` + 

poly.sa.habitat [year == 2017, mpb_m_8_year_area_ha_2017 := `X2017_M_survey2017` + `X2016_M_survey2017` + `X2015_M_survey2017` + `X2014_M_survey2017` + `X2013_M_survey2017` + `X2012_M_survey2017` + `X2011_M_survey2017` + `X2010_M_survey2017`]
poly.sa.habitat [year == 2017, mpb_s_8_year_area_ha_2017 :=  `X2017_S_survey2017` +  `X2016_S_survey2017` +   `X2012_S_survey2017` +   `X2010_S_survey2017`] # no data  `X2015_S_survey2017` + `X2014_S_survey2017` + `X2013_S_survey2017` + `X2011_S_survey2017` +
poly.sa.habitat [year == 2017, mpb_vs_8_year_area_ha_2017 :=  `X2011_V_survey2017` ] # no data  `X2017_V_survey2017` + `X2016_V_survey2017` + `X2015_V_survey2017` +  `X2014_V_survey2017` + `X2013_V_survey2017` + `X2012_V_survey2017`+ `X2010_V_survey2017`

poly.sa.habitat [year == 2018, mpb_m_8_year_area_ha_2018 := `X2018_M_survey2018` + `X2017_M_survey2018` + `X2016_M_survey2018` + `X2015_M_survey2018` + `X2014_M_survey2018` + `X2013_M_survey2018` + `X2012_M_survey2018` + `X2011_M_survey2018`]
poly.sa.habitat [year == 2018, mpb_s_8_year_area_ha_2018 :=  `X2018_S_survey2018` + `X2017_S_survey2018` + `X2016_S_survey2018` + `X2012_S_survey2018`] # no data  `X2015_S_survey2018` + `X2014_S_survey2018` + `X2013_S_survey2018` + `X2011_S_survey2018` +  
poly.sa.habitat [year == 2018, mpb_vs_8_year_area_ha_2018 :=   `X2011_V_survey2018` ] # no data  `X2018_V_survey2018` + `X2017_V_survey2018` + `X2016_V_survey2018` + `X2015_V_survey2018` + `X2014_V_survey2018` +  `X2013_V_survey2018` + `X2012_V_survey2018`+


poly.sa.habitat [year == 2014, mpb_m_15_year_area_ha_2014 := `X2014_M_survey2014` + `X2013_M_survey2014` + `X2012_M_survey2014` + `X2011_M_survey2014` + `X2010_M_survey2014` + `X2009_M_survey2014` + `X2008_M_survey2014` + `X2007_M_survey2014` + `X2006_M_survey2014` + `X2005_M_survey2014` + `X2004_M_survey2014` + `X2003_M_survey2014` + `X2002_M_survey2014` + `X2001_M_survey2014` + `X2000_M_survey2014`]
poly.sa.habitat [year == 2014, mpb_s_15_year_area_ha_2014 :=   `X2012_S_survey2014` + `X2010_S_survey2014` + `X2009_S_survey2014` + `X2008_S_survey2014` + `X2007_S_survey2014` + `X2006_S_survey2014` + `X2005_S_survey2014` + `X2004_S_survey2014` + `X2003_S_survey2014` + `X2002_S_survey2014` + `X2001_S_survey2014` + `X2000_S_survey2014`] # no data `X2014_S_survey2014` + `X2013_S_survey2014` + `X2011_S_survey2014` + 
poly.sa.habitat [year == 2014, mpb_vs_15_year_area_ha_2014 :=   `X2011_V_survey2014` +  `X2009_V_survey2014` + `X2008_V_survey2014` + `X2007_V_survey2014` + `X2006_V_survey2014` + `X2005_V_survey2014` + `X2004_V_survey2014` + `X2003_V_survey2014` ]  # no data `X2014_V_survey2014` +  `X2013_V_survey2014` + `X2012_V_survey2014` + `X2010_V_survey2014` + `X2002_V_survey2014` + `X2001_V_survey2014` +   `X2000_V_survey2014`

poly.sa.habitat [year == 2015, mpb_m_15_year_area_ha_2015 := `X2015_M_survey2015` + `X2014_M_survey2015` + `X2013_M_survey2015` + `X2012_M_survey2015` + `X2011_M_survey2015` + `X2010_M_survey2015` + `X2009_M_survey2015` + `X2008_M_survey2015` + `X2007_M_survey2015` + `X2006_M_survey2015` + `X2005_M_survey2015` + `X2004_M_survey2015` + `X2003_M_survey2015` + `X2002_M_survey2015` + `X2001_M_survey2015`]
poly.sa.habitat [year == 2015, mpb_s_15_year_area_ha_2015 :=   `X2012_S_survey2015` + `X2010_S_survey2015` + `X2009_S_survey2015` + `X2008_S_survey2015` + `X2007_S_survey2015` + `X2006_S_survey2015` + `X2005_S_survey2015` + `X2004_S_survey2015` + `X2003_S_survey2015` + `X2002_S_survey2015` + `X2001_S_survey2015`]  # no data `X2015_S_survey2015` + `X2014_S_survey2015` +  `X2013_S_survey2015` + `X2011_S_survey2015` + 
poly.sa.habitat [year == 2015, mpb_vs_15_year_area_ha_2015 :=   `X2011_V_survey2015` + `X2009_V_survey2015` + `X2008_V_survey2015` + `X2007_V_survey2015` + `X2006_V_survey2015` + `X2005_V_survey2015` + `X2004_V_survey2015` + `X2003_V_survey2015` ] # no data `X2015_V_survey2015` + `X2014_V_survey2015` + `X2013_V_survey2015` +   `X2012_V_survey2015` + `X2010_V_survey2015` + `X2002_V_survey2015` +  `X2001_V_survey2015`

poly.sa.habitat [year == 2016, mpb_m_15_year_area_ha_2016 := `X2016_M_survey2016` + `X2015_M_survey2016` + `X2014_M_survey2016` + `X2013_M_survey2016` + `X2012_M_survey2016` + `X2011_M_survey2016` + `X2010_M_survey2016` + `X2009_M_survey2016` + `X2008_M_survey2016` + `X2007_M_survey2016` + `X2006_M_survey2016` + `X2005_M_survey2016` + `X2004_M_survey2016` + `X2003_M_survey2016` + `X2002_M_survey2016`]
poly.sa.habitat [year == 2016, mpb_s_15_year_area_ha_2016 := `X2016_S_survey2016` + `X2012_S_survey2016` + `X2010_S_survey2016` + `X2009_S_survey2016` + `X2008_S_survey2016` + `X2007_S_survey2016` + `X2006_S_survey2016` + `X2005_S_survey2016` + `X2004_S_survey2016` + `X2003_S_survey2016` + `X2002_S_survey2016`] # no data `X2015_S_survey2016` + `X2014_S_survey2016` + `X2013_S_survey2016` + `X2011_S_survey2016` + 
poly.sa.habitat [year == 2016, mpb_vs_15_year_area_ha_2016 :=   `X2011_V_survey2016` +  `X2009_V_survey2016` + `X2008_V_survey2016` + `X2007_V_survey2016` + `X2006_V_survey2016` + `X2005_V_survey2016` + `X2004_V_survey2016` + `X2003_V_survey2016`] # no data `X2016_V_survey2016` +  `X2015_V_survey2016` +  `X2014_V_survey2016` + `X2013_V_survey2016` + `X2012_V_survey2016` + `X2010_V_survey2016`  + `X2002_V_survey2016`

poly.sa.habitat [year == 2017, mpb_m_15_year_area_ha_2017 := `X2017_M_survey2017` + `X2016_M_survey2017` + `X2015_M_survey2017` + `X2014_M_survey2017` + `X2013_M_survey2017` + `X2012_M_survey2017` + `X2011_M_survey2017` + `X2010_M_survey2017` + `X2009_M_survey2017` + `X2008_M_survey2017` + `X2007_M_survey2017` + `X2006_M_survey2017` + `X2005_M_survey2017` + `X2004_M_survey2017` + `X2003_M_survey2017`]
poly.sa.habitat [year == 2017, mpb_s_15_year_area_ha_2017 := `X2017_S_survey2017` + `X2016_S_survey2017` +  `X2012_S_survey2017` +  `X2010_S_survey2017` + `X2009_S_survey2017` + `X2008_S_survey2017` + `X2007_S_survey2017` + `X2006_S_survey2017` + `X2005_S_survey2017` + `X2004_S_survey2017` + `X2003_S_survey2017`] # no data `X2015_S_survey2017` + `X2014_S_survey2017` + `X2013_S_survey2017` + `X2011_S_survey2017` +
poly.sa.habitat [year == 2017, mpb_vs_15_year_area_ha_2017 :=   `X2011_V_survey2017` + `X2009_V_survey2017` + `X2008_V_survey2017` + `X2007_V_survey2017` + `X2006_V_survey2017` + `X2005_V_survey2017` + `X2004_V_survey2017` + `X2003_V_survey2017`] # no data: `X2017_V_survey2017` + `X2016_V_survey2017` +  `X2015_V_survey2017` +  `X2014_V_survey2017` + `X2013_V_survey2017` +   `X2012_V_survey2017` + `X2010_V_survey2017` + 

poly.sa.habitat [year == 2018, mpb_m_15_year_area_ha_2018 := `X2018_M_survey2018` +  `X2017_M_survey2018` + `X2016_M_survey2018` + `X2015_M_survey2018` + `X2014_M_survey2018` + `X2013_M_survey2018` + `X2012_M_survey2018` + `X2011_M_survey2018` + `X2010_M_survey2018` + `X2009_M_survey2018` + `X2008_M_survey2018` + `X2007_M_survey2018` + `X2006_M_survey2018` + `X2005_M_survey2018` + `X2004_M_survey2018`]
poly.sa.habitat [year == 2018, mpb_s_15_year_area_ha_2018 := `X2018_S_survey2018` + `X2017_S_survey2018` + `X2016_S_survey2018` +  `X2012_S_survey2018` + `X2010_S_survey2018` + `X2009_S_survey2018` + `X2008_S_survey2018` + `X2007_S_survey2018` + `X2006_S_survey2018` + `X2005_S_survey2018` + `X2004_S_survey2018`] # no data: `X2015_S_survey2018` + `X2014_S_survey2018` +  `X2013_S_survey2018` + `X2011_S_survey2018` +
poly.sa.habitat [year == 2018, mpb_vs_15_year_area_ha_2018 := `X2011_V_survey2018` +  `X2009_V_survey2018` + `X2008_V_survey2018` + `X2007_V_survey2018` + `X2006_V_survey2018` + `X2005_V_survey2018` + `X2004_V_survey2018`] # no data: `X2018_V_survey2018` +  `X2017_V_survey2018` + `X2016_V_survey2018` + `X2015_V_survey2018` +  `X2014_V_survey2018` + `X2013_V_survey2018` +  `X2012_V_survey2018` + `X2010_V_survey2018` +

poly.sa.habitat[is.na(poly.sa.habitat)] <- 0

poly.sa.habitat [, mpb_m_8_year_area_ha :=   `mpb_m_8_year_area_ha_2014` + `mpb_m_8_year_area_ha_2015` + `mpb_m_8_year_area_ha_2016` +  `mpb_m_8_year_area_ha_2017` + `mpb_m_8_year_area_ha_2018`]
poly.sa.habitat [, mpb_s_8_year_area_ha :=   `mpb_s_8_year_area_ha_2014` + `mpb_s_8_year_area_ha_2015` + `mpb_s_8_year_area_ha_2016` +  `mpb_s_8_year_area_ha_2017` + `mpb_s_8_year_area_ha_2018`]
poly.sa.habitat [, mpb_vs_8_year_area_ha :=   `mpb_vs_8_year_area_ha_2014` + `mpb_vs_8_year_area_ha_2015` + `mpb_vs_8_year_area_ha_2016` +  `mpb_vs_8_year_area_ha_2017` + `mpb_vs_8_year_area_ha_2018`]

poly.sa.habitat [, mpb_all_8_year_area_ha :=   `mpb_m_8_year_area_ha` + `mpb_s_8_year_area_ha` + `mpb_vs_8_year_area_ha`]

poly.sa.habitat [, mpb_m_15_year_area_ha :=   `mpb_m_15_year_area_ha_2014` + `mpb_m_15_year_area_ha_2015` + `mpb_m_15_year_area_ha_2016` +  `mpb_m_15_year_area_ha_2017` + `mpb_m_15_year_area_ha_2018`]
poly.sa.habitat [, mpb_s_15_year_area_ha :=   `mpb_s_15_year_area_ha_2014` + `mpb_s_15_year_area_ha_2015` + `mpb_s_15_year_area_ha_2016` +  `mpb_s_15_year_area_ha_2017` + `mpb_s_15_year_area_ha_2018`]
poly.sa.habitat [, mpb_vs_15_year_area_ha :=   `mpb_vs_15_year_area_ha_2014` + `mpb_vs_15_year_area_ha_2015` + `mpb_vs_15_year_area_ha_2016` +  `mpb_vs_15_year_area_ha_2017` + `mpb_vs_15_year_area_ha_2018`]

poly.sa.habitat [, mpb_all_15_year_area_ha :=   `mpb_m_15_year_area_ha` + `mpb_s_15_year_area_ha` + `mpb_vs_15_year_area_ha`]


poly.sa.habitat.sum <- poly.sa.habitat %>%
                        select (unique_id,	id,	year,	name_sa,	area_ha,	cut_8_year_area_ha,	cut_15_year_area_ha,	cut_25_year_area_ha,	mpb_all_8_year_area_ha,	mpb_all_15_year_area_ha,	mpb_m_8_year_area_ha,	mpb_s_8_year_area_ha,	mpb_vs_8_year_area_ha,	mpb_m_15_year_area_ha,	mpb_s_15_year_area_ha,	mpb_vs_15_year_area_ha)

# save the data
write.csv (poly.sa.habitat, here ("./data/study_area_habitat.csv"))
write.csv (poly.sa.habitat.sum, here ("./data/study_area_habitat_sum.csv"))

```

### Estimating Population Parameters
- currently using data provided in annual report; not divided into type (human, etc.)
- dependent variables are currently:
    - annual adult female survival rate; all mortality types
    - change in adult annual survival rate; all mortality types
    - calves/100 cows and change in calves/100 cows (indicator of recruitment)

```{r population params table, echo = F, message = F, warning = F, eval = T}
table.pop.data = data.table(
  unique_id = c ("Big Creek2015", "Big Creek2016", "Big Creek2017", "Big Creek2018", "Bonaparte2015",
                 "Bonaparte2016", "Bonaparte2017", "Bonaparte2018", "Entiako2014", "Entiako2015", 
                 "Entiako2016", "Entiako2017", "Entiako2018", "John Prince Research Forest2015",
                 "John Prince Research Forest2016", "John Prince Research Forest2017",
                 "John Prince Research Forest2018", "Prince George South2015", 
                 "Prince George South2016", "Prince George South2017", "Prince George South2018"),
  af_survival_rate_all = c (0.923, 0.862, 0.897, 0.901, 0.912, 0.871, 0.906, 0.977, 0.970, 0.913,
                            0.799, 0.775, 0.817, 1, 0.923, 0.949, 0.951, 0.875, 0.842, 0.946, 0.794),
  af_survival_rate_all_95CI = c (0.085, 0.103, 0.143, 0.092, 0.130, 0.092, 0.124, 0.045, NA, 0.081, 
                                 0.111, 0.168, 0.133, 0, 0.084, 0.110, 0.066, 0.162, 0.118, 0.151, 
                                 0.122),
  calves_100cows = c(37, 33, 27, 32, 25, 26, 16, 38, NA, NA, 14, 9, 15, 8, 17, 40, 37, 39, 27, 40,
                     26),
  density = c(170, NA, NA, 220, 296, NA, NA, 254, NA, 267, NA, NA, 217, 770, NA, NA, 490, 630, NA, 
              NA, 400),
  density_95CI = c(39, NA, NA, 38, 18, NA, NA, 41, NA, 45, NA, NA, 46, 93, NA, NA, 84, 102, NA, NA, 
                   78)
)

kable (table.pop.data,
       caption = "<b>Table 1. Moose Population Parameter Estimates.<b>",
       col.names = c("Study Area",
                     "Adult Female Survival Rate",
                      "Adult Female Survival Rate 95%CI",
                      "Calves/100 Cows",
                     "Density",
                     "Density 95%CI")) %>%
  kable_styling (position = "left",
                 bootstrap_options = c("striped", "hover"),
                 fixed_thead = T,
                 full_width = F,
                 font_size = 11) 
```

```{r, link pop params to habitat covariates, echo = F, message = F, warning = F, eval = F, include = F}

# link pop table to habitat table
setkey (table.pop.data, unique_id)
setkey (poly.sa.habitat.sum, unique_id)
data <- merge (table.pop.data, poly.sa.habitat.sum, all.x = TRUE)

# convert areas to proportion of study area
data [, prop_cut_8_year_area_ha := cut_8_year_area_ha / area_ha]
data [, prop_cut_15_year_area_ha := cut_15_year_area_ha / area_ha]
data [, prop_cut_25_year_area_ha := cut_25_year_area_ha / area_ha]
data [, prop_mpb_m_8_year_area_ha := mpb_m_8_year_area_ha / area_ha]
data [, prop_mpb_m_15_year_area_ha := mpb_m_15_year_area_ha / area_ha]
data [, prop_mpb_s_8_year_area_ha := mpb_s_8_year_area_ha / area_ha]
data [, prop_mpb_s_15_year_area_ha := mpb_s_15_year_area_ha / area_ha]
data [, prop_mpb_vs_8_year_area_ha := mpb_vs_8_year_area_ha / area_ha]
data [, prop_mpb_vs_15_year_area_ha := mpb_vs_15_year_area_ha / area_ha]
data [, prop_mpb_all_8_year_area_ha := mpb_all_8_year_area_ha / area_ha]
data [, prop_mpb_all_15_year_area_ha := mpb_all_15_year_area_ha / area_ha]

# get change in values for comapring rates of change
data [, delta_af_survival_rate_all := af_survival_rate_all - shift (af_survival_rate_all, n = 1, type = "lag"), by = name_sa]
data [, delta_calves_100cows := calves_100cows - shift (calves_100cows, n = 1, type = "lag"), by = name_sa]
data [, delta_cut_8_year_area_ha := cut_8_year_area_ha - shift (cut_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_cut_15_year_area_ha := cut_15_year_area_ha - shift (cut_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_cut_25_year_area_ha := cut_25_year_area_ha - shift (cut_25_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_m_8_year_area_ha := mpb_m_8_year_area_ha - shift (mpb_m_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_s_8_year_area_ha := mpb_s_8_year_area_ha - shift (mpb_s_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_vs_8_year_area_ha := mpb_vs_8_year_area_ha - shift (mpb_vs_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_all_8_year_area_ha := mpb_all_8_year_area_ha - shift (mpb_all_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_m_15_year_area_ha := mpb_m_15_year_area_ha - shift (mpb_m_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_s_15_year_area_ha := mpb_s_15_year_area_ha - shift (mpb_s_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_vs_15_year_area_ha := mpb_vs_15_year_area_ha - shift (mpb_vs_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_mpb_all_15_year_area_ha := mpb_all_15_year_area_ha - shift (mpb_all_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_cut_8_year_area_ha := prop_cut_8_year_area_ha - shift (prop_cut_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_cut_15_year_area_ha := prop_cut_15_year_area_ha - shift (prop_cut_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_cut_25_year_area_ha := prop_cut_25_year_area_ha - shift (prop_cut_25_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_m_8_year_area_ha := prop_mpb_m_8_year_area_ha - shift (prop_mpb_m_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_m_15_year_area_ha := prop_mpb_m_15_year_area_ha - shift (prop_mpb_m_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_s_8_year_area_ha := prop_mpb_s_8_year_area_ha - shift (prop_mpb_s_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_s_15_year_area_ha := prop_mpb_s_15_year_area_ha - shift (prop_mpb_s_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_vs_8_year_area_ha := prop_mpb_vs_8_year_area_ha - shift (prop_mpb_vs_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_vs_15_year_area_ha := prop_mpb_vs_15_year_area_ha - shift (prop_mpb_vs_15_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_all_8_year_area_ha := prop_mpb_all_8_year_area_ha - shift (prop_mpb_all_8_year_area_ha, n = 1, type = "lag"), by = name_sa]
data [, delta_prop_mpb_all_15_year_area_ha := prop_mpb_all_15_year_area_ha - shift (prop_mpb_all_15_year_area_ha, n = 1, type = "lag"), by = name_sa]

data [is.na (data)] <- 0

write.csv (data, here ("./data/data.csv"))
```

## Preliminary Results
Here I provide a preliminary description of the relationship between landscape disturbance and estimates of moose adult female survival or calves/100 cows. Entiako was typically the largest study area, at approximately 1,000,000 ha, and Bonaparte was typically the smallest study area, at approximately 300,000 ha (Table 2). The JPRF study area had typically the highest amount of less than 8 year old cutblocks, and Prince George South had the highest amount of less than 15 year old cutblocks. The Big Creek study area typically had the most less than 8 year old MPB disturbed forest stands and Prince George South had the highest amount of less than 15 year MPB disturbed forest stands.

```{r, table of disturbance by study area, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
data.disturb.summary <- data %>%
                          select (name_sa, year, area_ha, cut_8_year_area_ha, cut_15_year_area_ha, 
                                  mpb_all_8_year_area_ha, mpb_all_15_year_area_ha)
data.disturb.summary <- data.disturb.summary [order (data.disturb.summary$name_sa, data.disturb.summary$year), ]

data.disturb.summary$cut_8_year_area_ha <- round (data.disturb.summary$cut_8_year_area_ha, digits = 0)
data.disturb.summary$cut_15_year_area_ha <- round (data.disturb.summary$cut_15_year_area_ha, digits = 0)
data.disturb.summary$mpb_all_8_year_area_ha <- round (data.disturb.summary$mpb_all_8_year_area_ha, digits = 0)
data.disturb.summary$mpb_all_15_year_area_ha <- round (data.disturb.summary$mpb_all_15_year_area_ha, digits = 0)
data.disturb.summary$area_ha <- round (data.disturb.summary$area_ha, digits = 0)
data.disturb.summary$year <- as.character(data.disturb.summary$year)
  
kable (data.disturb.summary,
       caption = "<b>Table 2. Summary of Cutblock and Mountain Pine Beetle Disturbance Area by Study Area and Year.<b>",
       col.names = c("Study Area",
                     "Year",
                     "Area (ha)",
                      "Area of Cutblocks Less Than 8 y.o. (ha)",
                      "Area of Cutblocks Less Than 15 y.o. (ha)",
                     "Area of MPB Disturbed Stands Less Than 8 y.o. (ha)",
                     "Area of MPB Disturbed Stands Less Than 15 y.o. (ha)"),
       format.args = list (big.mark = ",")
       ) %>%
  kable_styling (position = "left",
                 bootstrap_options = c("striped", "hover"),
                 fixed_thead = T,
                 full_width = F,
                 font_size = 11) 
```

Trends in adult female moose survival rates varied across the study areas from 2015 to 2018 (Fig. 1). Survival was increasing in the Bonaparte, Entiako adn JPRF study areas, and decreasing in the Big Creek and Prince George South study areas. 

```{r, trend in survival over time, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = year, 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 1. Trend in Adult Female Moose Survival Rate by Study Area Over Time",
        x = "Year",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

Trends in calf/100 cow ratios varied across the study areas from 2015 to 2018 (Fig. 2). Calf/100 cow ratios were increasing in the Bonaparte study area, decreasing in the Entiako, JPRF and Prince Goerge SOtuh study areas, and flat in the Big Creek study area. 

```{r, trend in calf/cow over time, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = year, 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 2. Trend in Calf/100 Cow Ratios by Study Area Over Time",
        x = "Year",
        y = "Calf/100 Cow Ratio") +
  theme (axis.text.x = element_text (angle = 45))
```

### Relationship Between Cutblocks and Adult Female Survival
There was no clear relationship between the amount of less than 8 year old cutblocks and adult female moose survival rates across the five study areas (Fig. 3). There was a negative relationship in the Entiako and JPRF study areas, a positive relationship in the Big Creek and Prince George South study areas and no relationship in the Bonaparte study area. 

```{r, area cut lt 8 yo x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (cut_8_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 3. Adult Female Moose Survival Rate by Total Area of Cutblocks Less Than 8 years 
old in the Study Area MCPs",
        x = "Total Area of Cutblocks Less Than 8 years old (km2)",
        y = "Adult Female Moose Survival Rate") 
```

Similarly, there was no clear relationship between the percent area less than 8 year old cutblocks and adult female moose survival rates across the five study areas (Fig. 4). There was a negative relationship in the Bonaparte and Entiako study areas, and a positive relationship in the Big Creek, JPRF and Prince George South study areas. 

```{r, perc area cut lt 8 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (prop_cut_8_year_area_ha*100), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 4. Adult Female Moose Survival Rate by Percent of Area that is 
Cutblocks Less Than 8 years old in the Study Area MCPs",
        x = "Percent of Area that is Cutblocks Less Than 8 years old",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

When considering the relationship between annual change in 8 year old cutblocks and annual change in adult female moose survival rates, there was no clear relationship across the five study areas (Fig. 5). There was a negative relationship in the Bonaparte and Entiako study areas, and a positive relationship in the Big Creek, JPRF and Prince George South study areas. 

```{r, delta perc area cut lt 8 years old x delta survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_cut_8_year_area_ha*100), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 5. Change in Adult Female Moose Survival Rate by Change in Percent 
of Area that is Cutblocks Less Than 8 years old in the Study Area MCPs",
        x = "Change in Percent of Area that is Cutblocks Less Than 8 years old",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

There was no clear relationship between the area of less than 15 year old cutblocks and adult female moose survival rates across the five study areas (Fig. 6). There was a negative relationship in the JPRF study area, a positive relationship in the Bonaparte and Entiako study areas and no relationship in the Big Creek and Prince George South study areas.

```{r, area cut lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (cut_15_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 6. Adult Female Moose Survival Rate by Total Area of Cutblocks <15 
years old in the Study Area MCPs",
        x = "Total Area of Cutblocks <15 years old (km2)",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```

There was generally a positive relationship between the percent area less than 15 year old cutblocks and adult female moose survival rates across the five study areas (Fig. 7). This did not support the hypothesis that more forestry disturbance increased mrotality risk for moose.  

```{r, perc cut lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (prop_cut_15_year_area_ha*100), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 7. Adult Female Moose Survival Rate by Percent of Area that is 
Cutblocks <15 years old in the Study Area MCPs",
        x = "Percent of Area that is Cutblocks <15 years old",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

There was also generally a positive relationship between the rate of change in area of cutblocks less than 15 years old and change in adult female moose survival rates across the five study areas (Fig. 8). This did not support the hypothesis that more forestry disturbance increased mrotality risk for moose. 

```{r, delta perc area cut lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_cut_15_year_area_ha*100), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 8. Change in Adult Female Moose Survival Rate by Change in Percent of 
Area that is Cutblocks <15 years old in the Study Area MCPs",
        x = "Change in Percent of Area that is Cutblocks <15 years old",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

### Relationship Between Cutblocks and Calf/Cow Ratios
There was no clear relationship between the area of less than 8 year old cutblocks and calf/cow ratios across the five study areas (Fig. 9). There was a positive relationship in the Entiako and JPRF study areas and no relationship in the Big Creek, Bonaparte and Prince George South study areas.

```{r, area cut lt 8 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (cut_8_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 9. Moose Calves/100 Cows Ratio by Total Area of Cutblocks 
less than 8 years old in the Study Area MCPs",
        x = "Total Area of Cutblocks Less Than 8 years old (km2)",
        y = "Calves/100 Cows") 
```

Similarly, there was no clear relationship between the change in percent area of less than 8 year old cutblocks and change in calf/cow ratios across the five study areas (Fig. 10). There was a positive relationship in the Big Creek, Entiako and Prince George South study areas and no relationship in the Bonaparte and JPRF study areas.

```{r, delta perc area cut lt 8 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_cut_8_year_area_ha*100), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 10. Change in Moose Calf/100 Cows Ratio by Change in Percent of Area 
that is Cutblocks Less Than 8 years old in the Study Area MCPs",
        x = "Change in Percent of Area that is Cutblocks Less Than 8 years old",
        y = "Change in Moose Calf/100 Cows Ratio") +
  theme (axis.text.x = element_text (angle = 45))
```

There was no clear relationship between the area of less than 15 year old cutblocks and calf/cow ratios across the five study areas (Fig. 11). There was a negative relationship in the Big Creek, Entiako and Prince George South study areas and a positive relationship in the Bonaparte and JPRF study areas.

```{r, area cut lt 15 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (cut_15_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 11. Moose Calf/100 Cows Ratio by Total Area of Cutblocks Less Than 15 
years old in the Study Area MCPs",
        x = "Total Area of Cutblocks Less Than 15 years old (km2)",
        y = "Calves/100 Cows") 
```

Similarly, there was no clear relationship between the change in percent area of less than 15 year old cutblocks and change in calf/cow ratios across the five study areas (Fig. 12). There was a negative relationship in the JPRF study area, a positive relationship in the Bonaparte and Prince George South study areas and no relationsip in the Big Creek and Entiako study areas. 

```{r, delta perc area cut lt 15 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_cut_15_year_area_ha*100), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 12. Change in Moose Calf/100 Cows Ratio by Change in Percent of Area 
that is Cutblocks <15 years old in the Study Area MCPs",
        x = "Change in Percent of Area that is Cutblocks <15 years old",
        y = "Change in Moose Calf/100 Cows Ratio") +
  theme (axis.text.x = element_text (angle = 45))
```

In general, there was no clear relationship between forest cutblock disturbance and moose population parameters in the five study areas over time. If anything, these preliminary results indicate an overall positive relationship between forestry cutblocks and moose survival, although this realtionship was not clear enough across parameters and study areas. 

### Relationship Between Mountain Pine Beetle Disturbance and Adult Female Survival
There was no clear relationship between the area of less than 8 year old MPB disturbed stands and adult female moose survival rates across the five study areas (Fig. 13). There was a positive relationship in the Entiako and JPRF study areas and no relationship in the Big Creek, Bonaparte and Prince George South study areas.

```{r, MPB lt 8 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_8_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 13. Adult Female Moose Survival Rate by Total Area of Mountain 
Pine Beetle (MPB) Disturbed Stands in the Study Area MCPs",
        x = "Total Area of MPB Disturbed Stands Less Than 8 years old (km2)",
        y = "Adult Female Moose Survival Rate") 
```

Similarly, there was no clear relationship between the percent area of less than 8 year old MPB disturbed stands and adult female moose survival rates across the five study areas (Fig. 14). There was a negative relationship in the Entiako and JPRF study areas, and no relationship in the Big Creek, Bonaparte and Prince George South study areas.

```{r, perc mpb lt 8 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (prop_mpb_all_8_year_area_ha*100), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 14. Adult Female Moose Survival Rate by Percent of Area that is 
Mountain Pine Beetle (MPB) Disturbed Stands Less Than 8 years old in the 
Study Area MCPs",
        x = "Percent of Area that is MPB Disturbed Stands Less Than 8 years old",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

There was no clear relationship between the change in percent area of less than 8 year old MPB disturbed stands and change in adult female moose survival rates across the five study areas (Fig. 15). There was a positive relationship in the Bonaparte study area, a negative relationship in Big Creek, Entiako and JPRF study areas and no relationship in the Prince George South study area.

```{r, delta perc mpb lt 8 years old x delta survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_mpb_all_8_year_area_ha*100), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 15. Change in Adult Female Moose Survival Rate by Change in Percent 
of Area that is Mountain Pine Beetle (MPB) Disturbed Stands Less Than 8 years 
old in the Study Area MCPs",
        x = "Change in Percent of Area that is Mountain Pine Beetle (MPB) Disturbed Stands 
Less Than 8 years old",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

There was generally a consistent, but weak negative relationship between the area of less than 15 year old MPB disturbed stands and adult female moose survival rates across the five study areas (Fig. 16). Adult female moose survival rates were lower in most study areas during periods when there was more area of less than 15 year old MPB disturbed stands. However, this relationship was not clear in the Prince George South study area.

```{r, area mpb lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_15_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 16. Adult Female Moose Survival Rate by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands Less Than 15 Years Old in the Study Area MCPs",
        x = "Total Area of MPB Disturbed Stands Less Than 15 Years Old (km2)",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```

There was no clear relationship between the percent area of less than 15 year old MPB disturbed stands and change in adult female moose survival rates across the five study areas (Fig. 17). There was a negative relationship in the Bonaparte and JPRF study areas and no relationship in the other study areas.

```{r, perc mpb lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (prop_mpb_all_15_year_area_ha*100), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 17. Adult Female Moose Survival Rate by Percent of Area that is 
Mountain Pine Beetle (MPB) Disturbed Stands <15 years old in the 
Study Area MCPs",
        x = "Percent of Area of MPB Disturbed Stands <15 years old",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

There was a consistent negative relationship between the change in percent area of less than 15 year old MPB disturbed stands and change in adult female moose survival rates across the five study areas (Fig. 18). As the area of less than 15 year old MPB disturbed stands increased, survival rate decreased.

```{r, delta perc area mpb lt 15 years old x delta survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_mpb_all_15_year_area_ha*100), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 18. Change in Adult Female Moose Survival Rate by Change in Percent 
of Area that is Mountain Pine Beetle (MPB) Disturbed Stands <15 years old 
in the Study Area MCPs",
        x = "Change in Percent of Area that is MPB Disturbed Stands  <15 years old",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45))
```

### Relationship Between Mountain Pine Beetle Disturbance and Calf/Cow Ratios
There was no clear relationship between the area of less than 8 year old MPB disturbed stands and the ratio of calves/100 cows across the five study areas (Fig. 19). There was a positive relationship in the Big Creek, Entiako and Prince George South study areas, a negative relationship in the JPRF study area and no relationship in the Bonaparte study area.

```{r, mpb lt 8 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_8_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 19. Moose Calf/100 Cows Ratio by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands Less Than 8 years old in the Study Area MCPs",
        x = "Total Area of Mountain Pine Beetle (MPB) Disturbed Stands Less Than 8 years old (km2)",
        y = "Calves/100 Cows") 
```

Similarly, there was no clear relationship between the percent area of less than 15 year old MPB disturbed stands and adult calf/cow ratios across the five study areas (Fig. 20). There was a negative relationship in the Big Creek and Prince George South study areas, positive in the JPRF study area and no relationship in the Bonaparte and Entiako study areas.

```{r, mpb lt 15 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_15_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 20. Moose Calf/100 Cows Ratio by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands Less Than 15 years old in the Study Area MCPs",
        x = "Total Area of Mountain Pine Beetle (MPB) Disturbed Stands Less Than 15 years old (km2)",
        y = "Calves/100 Cows") 
```

There was no clear relationship between the change in area of less than 8 year old MPB disturbed stands and the change in the ratio of calves/100 cows across the five study areas (Fig. 21). There was a positive relationship in the Big Creek, Entiako and Prince George South study areas, a negative relationship in the JPRF study area and no relationship in the Bonaparte study area.

```{r, delta perc mpb lt 8 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_mpb_all_8_year_area_ha*100), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 21. Change in Moose Calf/100 Cows Ratio by Change in Percent of Area 
that is Mountain Pine Beetle (MPB) Disturbed Stands Less Than 8 years old in the Study Area MCPs",
        x = "Change in Percent of Area that is Mountain Pine Beetle (MPB) 
Disturbed Stands Less Than 8 years old",
        y = "Change in Moose Calf/100 Cows Ratio") +
  theme (axis.text.x = element_text (angle = 45))
```

Finally, there was no clear relationship between the change in area of less than 15 year old MPB disturbed stands and the change in the ratio of calves/100 cows across the five study areas (Fig. 22). There was a positive relationship in the Big Creek and Bonaparte study areas, a negative relationship in the Prince George South study area and no relationship in the Entiako and JPRF study areas.

```{r, delta perc mpb lt 15 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_prop_mpb_all_15_year_area_ha*100), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 22. Change in Moose Calf/100 Cows Ratio by Change in Percent of Area 
that is Mountain Pine Beetle (MPB) Disturbed Stands Less Than 15 years old in 
the Study Area MCPs",
        x = "Change in Percent of Area that is Mountain Pine Beetle (MPB) Disturbed Stands 
Less Than 15 years old",
        y = "Change in Moose Calf/100 Cows Ratio") +
  theme (axis.text.x = element_text (angle = 45))
```

### Cumulative Disturbance 
I considered the relationship between adult female moose survival rate and MPB disturbed stands plus cutblocks (i.e., cumulative disturbance). Note that care should be taken in interpreting these cumulative disturbance results, as they don't account for spatial overlap of disturabnce types, and thus 'double-count' disturbances where they overlap. Nevertheless, here they likely provide some indication of whether both disturbance types together might be related to moose populaiton parameter estimates.

I found no clear relationship between adult female moose survival rate and cumulative disturbance less than 8 (Fig. 23) or 15 (Fig. 24) years old.

```{r, area lt 8 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_8_year_area_ha*0.01) + (cut_8_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 23. Adult Female Moose Survival Rate by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands and Cutblocks Less Than 8 Years Old in the 
Study Area MCPs",
        x = "Total Area of Disturbed Stands Less Than 8 Years Old (km2)",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```

```{r, area lt 15 years old x survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_15_year_area_ha*0.01) + (cut_15_year_area_ha*0.01), 
                   y = af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 24. Adult Female Moose Survival Rate by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands and Cutblocks Less Than 15 Years Old in the 
Study Area MCPs",
        x = "Total Area of Disturbed Stands Less Than 15 Years Old (km2)",
        y = "Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```

I also found no clear relationship between change in adult female moose survival rate and change cumulative disturbance less than 8 (Fig. 25) or 15 (Fig. 26) years old.

```{r, delta area lt 8 years old x delta survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_mpb_all_8_year_area_ha*0.01) + (delta_cut_8_year_area_ha*0.01), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 25. Change in Adult Female Moose Survival Rate by Change in Total 
Area of Mountain Pine Beetle (MPB) Disturbed Stands and Cutblocks Less 
Than 8 Years Old in the Study Area MCPs",
        x = "Change in Total Area of Disturbed Stands Less Than 8 Years Old (km2)",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```


```{r, delta area lt 15 years old x delta survival rate, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_mpb_all_15_year_area_ha*0.01) + (delta_cut_15_year_area_ha*0.01), 
                   y = delta_af_survival_rate_all)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 26. Change in Adult Female Moose Survival Rate by Change in Total 
Area of Mountain Pine Beetle (MPB) Disturbed Stands and Cutblocks Less 
Than 15 Years Old in the Study Area MCPs",
        x = "Change in Total Area of Disturbed Stands Less Than 15 Years Old (km2)",
        y = "Change in Adult Female Moose Survival Rate") +
  theme (axis.text.x = element_text (angle = 45)) 
```

I considered the relationship between cow/calf ratios and MPB disturbed stands plus cutblocks (i.e., cumulative disturbance). I found no clear relationship between adult female moose survival rate and cumulative disturbance less than 8 (Fig. 27) or 15 (Fig. 28) years old.

```{r, area lt 8 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_8_year_area_ha*0.01) + (cut_8_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 27. Calf/100 Cows Ratios by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands and Cutblocks Less Than 8 Years Old in the 
Study Area MCPs",
        x = "Total Area of Disturbed Stands Less Than 8 Years Old (km2)",
        y = "Calf/100 Cow Ratio") +
  theme (axis.text.x = element_text (angle = 45)) 
```

```{r, area lt 15 years old x cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (mpb_all_15_year_area_ha*0.01) + (cut_15_year_area_ha*0.01), 
                   y = calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 28. Calf/100 Cows Ratios by Total Area of Mountain Pine 
Beetle (MPB) Disturbed Stands and Cutblocks Less Than 15 Years Old in the 
Study Area MCPs",
        x = "Total Area of Disturbed Stands Less Than 15 Years Old (km2)",
        y = "Calf/100 Cow Ratio") +
  theme (axis.text.x = element_text (angle = 45)) 
```

I also found no clear relationship between change in adult female moose survival rate and change cumulative disturbance less than 8 (Fig. 29) or 15 (Fig. 30) years old.

```{r, delta area lt 8 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_mpb_all_8_year_area_ha*0.01) + (delta_cut_8_year_area_ha*0.01), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 29. Change in Adult Female Moose Survival Rate by Change in Total 
Area of Mountain Pine Beetle (MPB) Disturbed Stands and Cutblocks Less 
Than 8 Years Old in the Study Area MCPs",
        x = "Change in Total Area of Disturbed Stands Less Than 8 Years Old (km2)",
        y = "Change in ACalf/100 Cow Ratio") +
  theme (axis.text.x = element_text (angle = 45)) 
```

```{r, delta area lt 15 years old x delta cow/calf, eval = T, echo = F, message = F, warning = F}
data <- read.csv (here ("./data/data.csv"))
levels (data$name_sa) [levels(data$name_sa) == "John Prince Research Forest"] <- "JPRF"
ggplot (data, aes (x = (delta_mpb_all_15_year_area_ha*0.01) + (delta_cut_15_year_area_ha*0.01), 
                   y = delta_calves_100cows)) +
  geom_point () +
  geom_smooth (method = lm) +
  facet_grid (.~name_sa, scales = "free") +
  labs (title = "Figure 30. Change in Adult Female Moose Survival Rate by Change in Total 
Area of Mountain Pine Beetle (MPB) Disturbed Stands and Cutblocks Less 
Than 15 Years Old in the Study Area MCPs",
        x = "Change in Total Area of Disturbed Stands Less Than 15 Years Old (km2)",
        y = "Change in Calf/100 Cow Ratio") +
  theme (axis.text.x = element_text (angle = 45)) 
```

## Preliminary Conclusions
Across the five study areas. there was no clear relationship between landscape disturbance, here defined as cutblocks and MBP distrubed stands, and moose population parameters, i.e., adult female survival and calves/100 cows delta area lt 8 years old x delta survival rate ratio. The exception may have been the relationship beetween less than 15 year old MPB disturbed stands and change in adult female survival rate (Fig. 16), as this was consistently negative across the five study areas. However, this may have just been a stastical fluke, as I examined the data in a multitude of ways, and there was a likelihood that at least one of these ways would show the realtionship I was looking for. 

Using the population parameter data here I cannot reject the null hypothesis of the LCH. However, this does not mean the LCH is not valid, but it is clearly diffcult to test at a 'study area scale' with only five study areas (i.e., n = 5) over a four year period. This justifies the need to implement long-term (i.e., multi-decade) monitoring programs to evaluate the effects of landscape change on wildlife.    